---
title: "Hello, World!"
---

### <br /> Check out my Data Analysis Website entirely made from R Markdowns. Work in Progress!!!

This is a short tutorial I made from the video tutorial on the dplyr package by Hadley Wickham. The link for the video tutorial:<br /> <a href="https://www.youtube.com/watch?v=8SGif63VW6E" target="_blank">Hadley Wickham's "dplyr" tutorial at useR 2014</a>

The "dplyr" package created by Hadley Wickham provides a set of easy-to-use functions to efficiently manipulate or tidy datasets in R. For this tutorial, we'll do some data operations on the flights dataset from the "nycflights13" package.

```{r, warning = F, message = F}
# Install the packages "dplyr" and "nycflights13" using the install.packages() function.

library(dplyr) # Loads dplyr package into the R session.
library(nycflights13) # Loads the flights dataset in R.
```

```{r, warning = F, message = F}
?flights # Check this out to get familiar with the data descriptions.
```

Let's check the number of observations, variables and variable types in the flights tibble/data frame.
```{r}
str(flights) # Displays the structure of flights tibble. 
```

In this tutorial we'll check out some common dplyr functions like filter(), select(), arrange(), mutate(), and summarize(). Let us begin the analysis by answering the following questions: <br />

### Flights going to San Francisco or Oakland:
```{r}
filter(flights, dest %in% c("SFO" ,"OAK")) # Using %in% operator to identify if the cities belongs to destination column. The %in% operator is similar to the == relational operator.
```

### Flights made in January:
```{r}
filter(flights, month == 1)
```

```{r}
# Another way to get flights in January if the month column was absent:
filter(flights, time_hour < "2013-02-01") # same result as previous. 
```

### Flights that departed between midnight and 5 AM:
```{r}
filter(flights, hour >= 0 & hour <= 5)
```

### Flights where arrival delay was twice as much as departure delay:
```{r}
filter(flights, arr_delay > (2*dep_delay))
```

### Display the departure delay and arrival delay columns:
```{r}
select(flights, dep_delay, arr_delay)
```

### Which flights were most delayed during departure?
```{r}
arrange(flights, desc(dep_delay))
```

### Compute speed in mph from time(in minutes) and distance(in miles). Which flight flew the fastest?
```{r}
mutate(flights, speed = distance/(air_time/60)) %>%
  arrange(desc(speed)) # The "%>%" (pipe operator) passes resulting data from the command before to the one after. So after adding the new columns, arrange() function will sort the resulting dataframe in descending order.
```

### Add a new variable that shows how much time was made up or lost in flight:
```{r}
mutate(flights, delta = dep_delay - arr_delay)
```

### Which destination has the highest average delays?
```{r}
flights %>%
  filter(!is.na(arr_delay)) %>%
  group_by(dest) %>%
  dplyr::summarize(Average_delay = mean(arr_delay), cases = n()) %>%
  filter(cases > 10) %>%
  arrange(desc(Average_delay))
```

### Which flights(i.e. carrier + flight) happen everyday? Where do they fly to?
```{r}
flights %>%
  group_by(carrier,flight,dest) %>%
  dplyr::summarize(n = n_distinct(time_hour)) %>%
  filter(n == 365)
```

### On average, how do delays vary over the course of the day? (Hint: hour + minute/60)
```{r}
flights %>%
  group_by(hour, minute) %>%
  dplyr::summarize(n = n(), mean = mean(na.omit(dep_delay)))
```

